---
title: "Variables"
output: html_document
date: "2026-06-15"
---
#Administrative 
```{r Packages} 
#Pacman load or open the packages that are needed for data and spatial management
# DATA PACKAGES
require(pacman)
pacman::p_load(dplyr, renv, readxl, tidyverse, pivottabler, stringr, purrr, reshape2, ggplot2, ggnewscale, viridis, pheatmap, hrbrthemes, gridExtra, ggpubr, scales,  purrr, GGally, stringdist,lubridate, ggpattern)


#MAPS PACKAGES
pacman::p_load(sf, sp, broom, dplyr, raster, tmap,  maps, boot,  gtools, ggmap, geosphere, stringr, foreign, ggspatial, mapproj, automap, leaflet, rgbif, terra, magrittr, stars,  spatialEco, geodist, exactextractr)



```


```{r Codes for filter - Complete}
#Here are the list of names for filter by state, using 2 code letters or complete names. Also, landcover categories

states_2code_letter <- c(
  "WA",
  "OR"
)

states_complete <- c(
  "Washington",
  "Oregon"
)


state_lookup <- as.data.frame(cbind(states_2code_letter,states_complete))


landcover_labels <- c(
  "0" = "No data",
  "1" = "Open water",
  "5" = "Evergreen forest",
  "6" = "Mixed forest",
  "8" = "Grassland",
  "10" = "Grassland2",
  "14" = "Wetlands",
  "15" = "Pasture/Agriculture",
  "16" = "Woody wetlands",
  "17" = "Urban",
  "18" = "Herbaceous wetlands"
)


#(The following list describes the display of land cover classification in the .tif file:
#> Value 1, Temperate or sub-polar needleleaf forest, RGB 0 61 0;
#> Value 2, Sub-polar taiga needleleaf forest, RGB 148 156 112;
#> Value 3, Tropical or sub-tropical broadleaf evergreen forest, RGB 0 99 0;
#> Value 4, Tropical or sub-tropical broadleaf deciduous forest, RGB 30 171 5;
#> Value 5, Temperate or sub-polar broadleaf deciduous forest, RGB 20 140 61;
#> Value 6, Mixed Forest, RGB 92 117 43;
#> Value 7, Tropical or sub-tropical shrubland, RGB 179 158 43;
#> Value 8, Temperate or sub-polar shrubland, RGB 179 138 51;
#> Value 9, Tropical or sub-tropical grassland, RGB 232 220 94;
#> Value 10, Temperate or sub-polar grassland, RGB 225 207 138;
#> Value 11, Sub-polar or polar shrubland-lichen-moss, RGB 156 117 84;
#> Value 12, Sub-polar or polar grassland-lichen-moss, RGB 186 212 143;
#> Value 13, Sub-polar or polar barren-lichen-moss, RGB 64 138 112;
#> Value 14, Wetland, RGB 107 163 138;
#> Value 15, Cropland, RGB 230 174 102;
#> Value 16, Barren lands, RGB 168 171 174;
#> Value 17, Urban, RGB 220 33 38;
#> Value 18, Water, RGB 76 112 163;
#> Value 19, Snow and Ice, RGB 255 250 255)


```


```{r State + County shp and Centriods - Complete}
#### States

shpStates <- read_sf("DATA/WA_OR_Shp/USA_Lines/geoBoundaries-USA-ADM1.shp") #change by the folder and file of use

shpStates1 <- shpStates %>%
   dplyr::filter(shapeName %in% states_complete)
#I Changed the shape file that I was using before (it was already croped to OR and WA but it didnt descern the two states) tried to file a simular file to what you use but it had a different column name.

shpStates0<- st_transform(shpStates1, 4326)

####Counties


shp <- read_sf("DATA/WA_OR_Shp/WA_OR_County_Lines/BLM_OR_County_Boundaries_Polygon_Hub.shp") #change by the folder and file of use

shp1 <- shp %>%
  mutate(shapename = case_when(
  str_starts(COBCODE, "WA") ~ "WA",
  str_starts(COBCODE, "OR") ~ "OR",)) #Creating a new column to easily identifty if its WA or OR alone

 shp_0 <- st_transform(shp1, 4326)
 shpstates0 <- st_transform(shpStates1, 4326)

##Filtrate and made centroids per county
 shp2 <- shp1 %>%
  st_point_on_surface() %>%   #centroides
  mutate(
lon = st_coordinates(.)[,1],
lat = st_coordinates(.)[,2]
)
shp2 <- shp2 %>%
  mutate(
    state_name = case_when(
      shapename == "OR" ~ "Oregon",
      shapename == "WA" ~ "Washington",
      ))

  shp2 <- shp2 %>%
   left_join(state_lookup, by=c("state_name" = "states_complete"))%>%
   mutate(
   state_name_upper = str_to_upper(state_name))
```


```{r planning regions - need to ask eval = FALSE }
###CT planning regions

#Identify the region per county
#https://geodata.ct.gov/datasets/ctmaps::ct-planning-regions-1/about
#CTreg https://data.ct.gov/w/8k4w-vrag/wqz6-rhce?cur=OwaGpCkVELK 


CTshp <- read_sf("folder/CT_Planning_Regions.shp")#change by the folder and file of use

#plot(CTshp)

CTreg_sel <- CTshp %>% dplyr::select(PlanningRe)

shpCTreg <- st_transform(shp2, st_crs(CTshp))

CTregshp <- st_join(shpCTreg, CTreg_sel, join = st_within)%>%
  dplyr::select(NAME_1, states_2code_letter, NAME_2,PlanningRe)%>%
  dplyr::filter(!is.na(PlanningRe))
  

Regions_values <- CTregshp$PlanningRe




```

#Variables
Here is the main variables used.
First is the access to the files. Then included a data cleaniing in case the names are different between datasets. Then, follows the filters and rearrengment of the data according with the needs, specially in cvs files. Finally the estimation or extraction needed (e.g. means, distances, extraction of raster data, etc.). Last steps is join dataset to the file of interest taking into account the order of join. 

```{r livestock markets - complete}
#### markets
#https://www.aphis.usda.gov/livestock-poultry-disease/traceability/approved-livestock-markets  

markets <- read.csv("DATA/Overview Map (2)_data.csv")

markets1 <- markets %>%
     dplyr::filter(`Prem.State.1` %in% states_2code_letter)

#filter by coordinates
#dont really need this one becuase I dont have any NA cords but im keeping it here just in case
markets1_clean <- markets1 |>
    dplyr::filter(
    !is.na(`Longitude..generated.`),
    !is.na(`Latitude..generated.`)
)


markets1_sf <- st_as_sf(
  markets1_clean,
  coords = c("Longitude..generated.", "Latitude..generated."),
  crs = 4326
  )

markets1_proj<- st_transform(markets1_sf, 4326)


# distance to the markets
shp2 <- st_transform(shp2, 4326)

dist_matrix <- st_distance(shp2, markets1_sf)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 


shp2$dist_markets_m <- as.numeric(min_dist)#added to the final database 


# number markets per county
shp_0 <- st_make_valid(shp_0)
shp2$n_markets <- lengths(st_intersects(shp_0, markets1_sf))
```


```{r Roads-complete}
##### Roads
#https://catalog.data.gov/dataset/tiger-line-shapefile-2016-nation-u-s-primary-roads-national-shapefile 

#We selected and filter specific roads 
Highways <- read_sf("DATA/Oregon_Washington_HW/BLM_OR_Ground_Transportation_GTRN_Highways_Line_Hub.shp")
roads_all <- filter(Highways, I_SIGN ==5|I_SIGN==84|I_SIGN==90|
                       ST_SIGN_1==97|ST_SIGN_2==97|US_SIGN_1==97|US_SIGN_2==97)
roads_all <- st_make_valid(roads_all)

roads_all_1<- st_transform(roads_all, 4326)


roads_clip <- st_intersection(roads_all_1, shp_0)


#distance to the roads
dist_matrix <- st_distance(shp2, roads_all_1)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 

shp2$dist_roads_m <- as.numeric(min_dist)#added to the final database
```


```{r Roads (just I-95 - emailed Zulma to ask why) eval = FALSE}
#dist roads I-95

roads3_1<- st_transform(roads3, 4326)

dist_matrix <- st_distance(shp2, roads3_1)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 

shp2$dist_roads_i95 <- as.numeric(min_dist)#added to the final database
```


```{r Prism Temp and Prec - complete}


# Temp
temp_file <- rast("DATA/Mean_Annual_Temp/prism_tmean_us_25m_2025.tif")#change by the folder and file of use#change by the folder and file of use

temp <-  project(temp_file, "EPSG:4326") 

temp_crop <- crop(temp, shpStates1)

temp_mask <- terra::mask(temp_crop, shpStates1)

#plot(temp_mask)

Tempvalue <- terra::extract(temp_mask, shp2)

shp2$Temperature <- Tempvalue[,2]#added to the final database 


# Precipitation
prec_file<- rast("DATA/Prec_DATA/prism_ppt_us_25m_2025.tif")#change by the folder and file of use

prec <-  project(prec_file, "EPSG:4326")

prec_crop <- crop(prec, shpStates1)

prec_mask <- terra::mask(prec_crop, shpStates1)

#plot(prec_mask)

precvalue <- terra::extract(prec_mask, shp2)

shp2$Precipitation <- precvalue[,2]#added to the final database
```


```{r LandCover - complete}
#### Landcover
#https://www.cec.org/north-american-environmental-atlas/land-cover-30m-2020/ 

land<- rast("DATA/usa_land_cover_2020v2_30m_tif/USA_NALCMS_landcover_2020v2_30m/data/USA_NALCMS_landcover_2020v2_30m.tif")#change by the folder and file of use

centroids_proj <- st_transform(shp2, crs(land))

#plot(land)

landvalue <- terra::extract(land, centroids_proj)
shp2$Landcover <- landvalue[,2]#added to the final database 

 shp2$Landcover_group <- dplyr::case_when(
     shp2$Landcover %in% c(
         "Temperate or Subpolar Needleaf Forest",
         "Sub-polar Taiga Needleleaf Forest",
         "Tropical or Sub-Tropical Broadleaf Evergreen Forest",
         "Tropical or sub-tropical broadleaf deciduous Forest",
         "Temperate or sub-polar broadleaf deciduous forest",
         "Mixed Forest"
     ) ~ "Forest",
     
     shp2$Landcover %in% c(
          "Tropical or Subtropical Shrubland",
         "Temperate or Subpolar Shrubland",
         "Tropical or Subtropical Grassland",
         "Temperate or Subpolar Grassland",
         "Sub-Polar or Polar Shrubland-Lichen-Moss",
         "Sub-Polar or Polar Grassland-Lichen-Moss",
         "Sub-Polar or Polar Barren-Lichen-moss",
         "Wetland",
         "Barren land",
         "Water",
         "Snow and Ice"
     ) ~ "Non_Forested",
    shp2$Landcover %in% c(
         "Cropland",
         "Urban and Built-up"
     ) ~ "Developed",
     
     TRUE ~ "No_data"
 )
```


```{r Deer - Cant do anything RN because waiting to hear back eval = FALSE }
##### Deer

#https://peerj.com/articles/8262/#supplemental-information

deer<- st_read("folder/QDMA_2001_2005.shp")#change by the folder and file of use

shpStates2 <- st_transform(shpStates1, st_crs(deer))

deer_crop <- st_intersection(deer, shpStates2)

#plot(deer_crop)

deer_sel <- deer_crop %>% dplyr::select(Per)

shpdeersel <- st_transform(shp2, st_crs(deer_sel))

deervalue <- st_join(shpdeersel, deer_sel, join = st_within)
per_values <- deervalue$Per

shp2$Deer_density <- deervalue$Per#added to the final database
```


```{r Livestock - Incomplete eval = FALSE}
##### livestock
#https://quickstats.nass.usda.gov/src/glossary.pdf 

livestock0<- read_csv("DATA/usa_land_cover_2020v2_30m_tif/USA_NALCMS_landcover_2020v2_30m/data/USA_NALCMS_landcover_2020v2_30m.tif")#change by the folder and file of use

#data cleaning
livestock0$Value <- livestock0$Value %>%
  gsub(",", "", .) %>%        # quitar comas
  gsub("\\(D\\)", NA, .) %>%  # reemplazar "(D)" por NA
  as.numeric()                 # convertir a número

livestock0$State <- str_to_title(livestock0$State)
livestock0$County <- str_to_title(livestock0$County)


livestock1<-livestock0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 5)
    )
  )%>%
  dplyr::select(Year, State,County, Value)

#reorganize database
LivestockYear <- livestock1 %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

#estimate means - Only using one year


shp2 <- shp2 %>%#added to the final database 
  left_join(
    LivestockYear %>%
      dplyr::select(State, County, total_cattle=Value),
    by = c("state_name" = "State",
           "COUNTY_NAM" = "County")
  )
```


```{r Beef - Complete}
##### beef
#https://quickstats.nass.usda.gov/src/glossary.pdf 

beef<- read_csv("DATA/Livestock/Beef_Total.csv")#change by the folder and file of use

#data cleaning
beef0<-beef%>%
  dplyr::filter(Domain== "TOTAL",)

beef0$Value <- beef0$Value %>%
  gsub(",", "", .) %>%        # remove comma
  gsub("\\(D\\)", NA, .) %>%  # replace "(D)" by NA
  as.numeric()                 # change to number

beef0$State <- str_to_title(beef0$State)
beef0$County <- str_to_title(beef0$County)


beef1<-beef0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 4)
    )
  )%>%mutate(County = str_to_title(County))%>%
  dplyr::select(Year, State,County, Value)%>%
    mutate(Value = replace(Value, !is.finite(Value), 0))
  

#reorganize database
beefYear <- beef1 %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

shp2 <- shp2 %>%
  mutate(
    state_name = str_to_title(str_trim(state_name)),
    COUNTY_NAM = str_to_title(str_trim(COUNTY_NAM))
  )

shp2 <- shp2 %>%#added to the final database 
  left_join(
    beefYear %>%
      dplyr::select(State, County, Beef_Cattle = Value),
    by = c("state_name" = "State",
           "COUNTY_NAM" = "County")
  )
#estimate means - don't need to do this because I am only using one year
```


```{r Dairy - Complete}
##### dairy
#https://quickstats.nass.usda.gov/src/glossary.pdf 

dairy0<- read_csv("DATA/Livestock/Dairy_Total.csv")#change by the folder and file of use

#data cleaning
dairy0$Value <- dairy0$Value %>%
  gsub(",", "", .) %>%        
  gsub("\\(D\\)", NA, .) %>%  
  as.numeric()                

dairy0$State <- str_to_title(dairy0$State)
dairy0$County <- str_to_title(dairy0$County)


dairy1<-dairy0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 5)
    )
  )%>%
  dplyr::select(Year, State,County, Value)%>%
    mutate(Value = replace(Value, !is.finite(Value), 0))

#reorganize database
dairyYear <- dairy1 %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


#estimate means - only using one year


shp2 <- shp2 %>%#added to the final database 
  left_join(
    dairyYear %>%
      dplyr::select(State, County, Dairy_Cattle = Value),
    by = c("state_name" = "State",
           "COUNTY_NAM" = "County")
  )
```


```{r Humans - Complete}
##### humans
#https://www.ers.usda.gov/data-products/rural-urban-continuum-codes 

humans<- read_xlsx("DATA/Population.xlsx")#change by the folder and file of use

#data cleaning
humans1 <- humans %>%
  filter(State %in% states_2code_letter) %>%
  mutate(
    County = str_remove(County_Name, " County"))%>%
  dplyr::select(State, County, Population_2020)
  

humans1 <- humans1 %>%
  mutate(
    COUNTY = ifelse(
      State == "CT",
      sub("^Capitol", "Capital", County),
      County
    )
  )

#Join the planning region for CT
#humans1_fixed <- humans1 %>%
  #mutate(
   # key6 = ifelse(State == "CT", substr(COUNTY, 1, 6), NA)
 # ) %>%
 # left_join(
 #   CTregshp %>%
 #     mutate(key6 = substr(PlanningRe, 1, 6)) %>%
 #     dplyr::select(key6, NAME_2),
 #   by = "key6"
 # ) %>%
  #mutate(
  #  COUNTY = ifelse(State == "CT" & !is.na(NAME_2), NAME_2, COUNTY)
  #) %>%
  #dplyr::select(-key6, -NAME_2)%>%
  #dplyr::select(State, COUNTY, Population_2020)


humans1_fixed <- humans1 %>%
  mutate(
    County = str_to_upper(str_trim(County)),
    State = str_to_upper(str_trim(State))
  )

shp2 <- shp2 %>%#added to the final database 
  left_join(
    humans1,
    by = c("COUNTY_NAM" = "County",
           "states_2code_letter" = "State")
  )
```


```{r Social index - Complete}
##### social index
#https://www.atsdr.cdc.gov/place-health/php/svi/svi-interactive-map.html


social<- read_csv("DATA/svi_interactive_map_USA.csv")#change by the folder and file of use

#data cleaning
social1 <- social %>%
  dplyr::filter(ST_ABBR %in% states_2code_letter) %>%
  dplyr::select(ST_ABBR, COUNTY, RPL_THEMES)%>%
  mutate(
    COUNTY = str_remove(COUNTY, " County"))

shp2 <- shp2 %>%#added to the final database 
  left_join(
    social1,
    by = c("COUNTY_NAM" = "COUNTY",
           "states_2code_letter" = "ST_ABBR")
  )
```


  )
```{r Join the planning region for CT eval = FALSE}
social1 <- social1 %>%
  mutate(
    COUNTY = ifelse(
      ST_ABBR == "CT",
      sub("^Capitol", "Capital", COUNTY),
      COUNTY
    )
  )
#Join the planning region for CT
social1_fixed <- social1 %>%
  mutate(
    key6 = ifelse(ST_ABBR == "CT", substr(COUNTY, 1, 6), NA)
  ) %>%
  left_join(
    CTregshp %>%
      mutate(key6 = substr(PlanningRe, 1, 6)) %>%
      dplyr::select(key6, NAME_2),
    by = "key6"
  ) %>%
  mutate(
    COUNTY = ifelse(ST_ABBR == "CT" & !is.na(NAME_2), NAME_2, COUNTY)
  ) %>%
  dplyr::select(-key6, -NAME_2)%>%
  dplyr::select(ST_ABBR, COUNTY, RPL_THEMES)


social1_fixed <- social1_fixed %>%
  mutate(
    COUNTY = str_to_upper(str_trim(COUNTY)),
    ST_ABBR = str_to_upper(str_trim(ST_ABBR))
  )


shp2 <- shp2 %>%#added to the final database 
  left_join(
    social1_fixed,
    by = c("states_2code_letter" = "ST_ABBR",
           "NAME_2_upper" = "COUNTY")
  )



```


