---
title: "Variables"
output: html_document
date: "2026-06-15"
---
#Packages
Pacman load or open the packages that are needed for data and spatial management

```{r}

# DATA PACKAGES
require(pacman)
pacman::p_load(dplyr, renv, readxl, tidyverse, pivottabler, stringr, purrr, reshape2, ggplot2, ggnewscale, viridis, pheatmap, hrbrthemes, gridExtra, ggpubr, scales,  purrr, GGally, stringdist,lubridate, ggpattern)


#MAPS PACKAGES
pacman::p_load(sf, sp, broom, dplyr, raster, tmap,  maps, boot,  gtools, ggmap, geosphere, stringr, foreign, ggspatial, mapproj, automap, leaflet, rgbif, terra, magrittr, stars,  spatialEco, geodist, exactextractr)



```

#codes for filter
Here are the list of names for filter by state, using 2 code letters or complete names. Also, landcover categories

```{r}


states_2code_letter <- c(
  "NY",
  "NJ",
  "WV",
  "VA",
  "MD",
  "DC",
  "PA",
  "NC",
  "DE",
  "CT",
  "RI"
)

states_complete <- c(
  "New York",
  "New Jersey",
  "West Virginia",
  "Virginia",
  "Maryland",
  "Washington DC",
  "Pennsylvania",
  "North Carolina",
  "Delaware",
  "Connecticut",
  "Rhode Island"
)

state_lookup <- as.data.frame(cbind(states_2code_letter,states_complete))


landcover_labels <- c(
  "0" = "No data",
  "1" = "Open water",
  "5" = "Evergreen forest",
  "6" = "Mixed forest",
  "8" = "Grassland",
  "10" = "Grassland2",
  "14" = "Wetlands",
  "15" = "Pasture/Agriculture",
  "16" = "Woody wetlands",
  "17" = "Urban",
  "18" = "Herbaceous wetlands"
)


#(The following list describes the display of land cover classification in the .tif file:
#> Value 1, Temperate or sub-polar needleleaf forest, RGB 0 61 0;
#> Value 2, Sub-polar taiga needleleaf forest, RGB 148 156 112;
#> Value 3, Tropical or sub-tropical broadleaf evergreen forest, RGB 0 99 0;
#> Value 4, Tropical or sub-tropical broadleaf deciduous forest, RGB 30 171 5;
#> Value 5, Temperate or sub-polar broadleaf deciduous forest, RGB 20 140 61;
#> Value 6, Mixed Forest, RGB 92 117 43;
#> Value 7, Tropical or sub-tropical shrubland, RGB 179 158 43;
#> Value 8, Temperate or sub-polar shrubland, RGB 179 138 51;
#> Value 9, Tropical or sub-tropical grassland, RGB 232 220 94;
#> Value 10, Temperate or sub-polar grassland, RGB 225 207 138;
#> Value 11, Sub-polar or polar shrubland-lichen-moss, RGB 156 117 84;
#> Value 12, Sub-polar or polar grassland-lichen-moss, RGB 186 212 143;
#> Value 13, Sub-polar or polar barren-lichen-moss, RGB 64 138 112;
#> Value 14, Wetland, RGB 107 163 138;
#> Value 15, Cropland, RGB 230 174 102;
#> Value 16, Barren lands, RGB 168 171 174;
#> Value 17, Urban, RGB 220 33 38;
#> Value 18, Water, RGB 76 112 163;
#> Value 19, Snow and Ice, RGB 255 250 255)


```

#Administrative shapefiles
Here are the main shapefile used for mapping.
It is made an initial filter for especific states

```{r}
#### States

shpStates <- read_sf("folder/USA_adm1.shp")#change by the folder and file of use

shpStates1 <- shpStates %>%
  dplyr::filter(NAME_1 %in% states_complete)

shpStates0<- st_transform(shpStates1, 5070)

####Counties


shp <- read_sf("folder/USA.shp")#change by the folder and file of use

shp1 <- shp %>%
  dplyr::filter(NAME_1 %in% states_complete)

crs(shp)

##Filtrate and made centroids per county
shp2 <- shp %>%
  dplyr::filter(NAME_1 %in% states_complete)%>%
  st_point_on_surface() %>%   #centroides
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  )

shp2 <- shp2 %>%
  left_join(state_lookup, by=c("NAME_1" = "states_complete"))%>%
  mutate(
    NAME_2_upper = str_to_upper(NAME_2),                # counties in capital letters
    #estados_2code_upper = str_to_upper(states_2code_letter)  #states in capital letters
  )

shp_0 <- st_transform(shp1, 5070)

###CT planning regions

#Identify the region per county
#https://geodata.ct.gov/datasets/ctmaps::ct-planning-regions-1/about
#CTreg https://data.ct.gov/w/8k4w-vrag/wqz6-rhce?cur=OwaGpCkVELK 


CTshp <- read_sf("folder/CT_Planning_Regions.shp")#change by the folder and file of use

#plot(CTshp)

CTreg_sel <- CTshp %>% dplyr::select(PlanningRe)

shpCTreg <- st_transform(shp2, st_crs(CTshp))

CTregshp <- st_join(shpCTreg, CTreg_sel, join = st_within)%>%
  dplyr::select(NAME_1, states_2code_letter, NAME_2,PlanningRe)%>%
  dplyr::filter(!is.na(PlanningRe))
  

Regions_values <- CTregshp$PlanningRe




```

#Variables
Here is the main variables used.
First is the access to the files. Then included a data cleaniing in case the names are different between datasets. Then, follows the filters and rearrengment of the data according with the needs, specially in cvs files. Finally the estimation or extraction needed (e.g. means, distances, extraction of raster data, etc.). Last steps is join dataset to the file of interest taking into account the order of join. 

```{r}
#### markets
#https://www.aphis.usda.gov/livestock-poultry-disease/traceability/approved-livestock-markets  

markets<- read_excel("folder/OverviewMap.xlsx")#change by the folder and file of use

markets1 <- markets %>%
  dplyr::filter(`Prem State...15` %in% states_2code_letter)
#NOT IN MARKET DATASET  #DC #CT #RI

#filter by coordinates
markets1_clean <- markets1 |>
  dplyr::filter(
    !is.na(`Longitud (generado)`),
    !is.na(`Latitud (generado)`)
  )

markets1_sf <- st_as_sf(
  markets1_clean,
  coords = c("Longitud (generado)", "Latitud (generado)"),
  crs = 4326   # WGS84
)

markets1_proj<- st_transform(markets1_sf, 5070)


# distance to the markets
dist_matrix <- st_distance(shp2, markets1_sf)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 


shp2$dist_markets_m <- as.numeric(min_dist)#added to the final database 


# number markets per county
shp2$n_markets <- lengths(st_intersects(shp1, markets1_sf))


##### Roads
#https://catalog.data.gov/dataset/tiger-line-shapefile-2016-nation-u-s-primary-roads-national-shapefile 

#We selected and filter specific roads 
roads1 <- st_read("folder/I77.shp")#change by the folder and file of use
roads2 <- st_read("folder/I81.shp")#change by the folder and file of use
roads3 <- st_read("folder/I95.shp")#change by the folder and file of use


roads_all <- rbind(roads1, roads2, roads3)
roads_all <- st_make_valid(roads_all)

roads_all_1<- st_transform(roads_all, 4326)


roads_all_proj<- st_transform(roads_all, 5070)
roads_clip <- st_intersection(roads_all_proj, shp_0)


#distance to the roads
dist_matrix <- st_distance(shp2, roads_all_1)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 

shp2$dist_roads_m <- as.numeric(min_dist)#added to the final database 



#dist roads I-95
roads3_1<- st_transform(roads3, 4326)

dist_matrix <- st_distance(shp2, roads3_1)#calculate euclidian distance 

min_dist <- apply(dist_matrix, 1, min)#take minimal distance 

shp2$dist_roads_i95 <- as.numeric(min_dist)#added to the final database 



##### Worldclim

# Temp
temp<- rast("folder/wc2.1_2.5m_bio_1.tif")#change by the folder and file of use

temp_crop <- crop(temp, shpStates1)

temp_mask <- terra::mask(temp_crop, shpStates1)

#plot(temp_mask)

Tempvalue <- terra::extract(temp_mask, shp2)

shp2$Temperature <- Tempvalue[,2]#added to the final database 


# Precipitation
prec<- rast("folder/wc2.1_2.5m_bio_12.tif")#change by the folder and file of use

prec_crop <- crop(prec, shpStates1)

prec_mask <- terra::mask(prec_crop, shpStates1)

#plot(prec_mask)

precvalue <- terra::extract(prec_mask, shp2)

shp2$Precipitation <- precvalue[,2]#added to the final database 



#### Landcover
#https://www.cec.org/north-american-environmental-atlas/land-cover-30m-2020/ 

land<- rast("folder/USA_NALCMS_landcover_2020v2_30m.tif")#change by the folder and file of use

centroids_proj <- st_transform(shp2, crs(land))

#plot(land)

landvalue <- terra::extract(land, centroids_proj)
shp2$Landcover <- landvalue[,2]#added to the final database 


shp2$Landcover_group <- dplyr::case_when(
  shp2$Landcover %in% c(1,2,3,4,5,6) ~ "Forest",
  shp2$Landcover %in% c(7,8,9,10,11, 12, 13,14,16, 18, 19) ~ "Non_Forested",#shrubland, grassland
  shp2$Landcover %in% c(15,17) ~ "Developed",
  TRUE ~ "No_data"
)



##### Deer

#https://peerj.com/articles/8262/#supplemental-information

deer<- st_read("folder/QDMA_2001_2005.shp")#change by the folder and file of use

shpStates2 <- st_transform(shpStates1, st_crs(deer))

deer_crop <- st_intersection(deer, shpStates2)

#plot(deer_crop)

deer_sel <- deer_crop %>% dplyr::select(Per)

shpdeersel <- st_transform(shp2, st_crs(deer_sel))

deervalue <- st_join(shpdeersel, deer_sel, join = st_within)
per_values <- deervalue$Per

shp2$Deer_density <- deervalue$Per#added to the final database 



##### livestock
#https://quickstats.nass.usda.gov/src/glossary.pdf 

livestock0<- read_csv("folder/CensusCattle_NASSstatequicks_2017_2022.csv")#change by the folder and file of use

#data cleaning
livestock0$Value <- livestock0$Value %>%
  gsub(",", "", .) %>%        # quitar comas
  gsub("\\(D\\)", NA, .) %>%  # reemplazar "(D)" por NA
  as.numeric()                 # convertir a número

livestock0$State <- str_to_title(livestock0$State)
livestock0$County <- str_to_title(livestock0$County)


livestock1<-livestock0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 5)
    )
  )%>%
  dplyr::select(Year, State,County, Value)

#reorganize database
LivestockYear <- livestock1 %>%
  pivot_wider(
    names_from = Year,
    values_from = Value)%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

#estimate means

LivestockYear$Cattle_mean<- LivestockYear$`2022` + LivestockYear$`2017`/ 2

shp2 <- shp2 %>%#added to the final database 
  left_join(
    LivestockYear %>%
      dplyr::select(State, County, `2017`, `2022`, Cattle_mean),
    by = c("NAME_1" = "State",
           "NAME_2" = "County")
  )


##### beef
#https://quickstats.nass.usda.gov/src/glossary.pdf 

beef<- read_csv("folder/CensusBeef_NASSstatequicks_2017_2022.csv")#change by the folder and file of use

#data cleaning
beef0<-beef%>%
  dplyr::filter(Domain== "TOTAL",)

beef0$Value <- beef0$Value %>%
  gsub(",", "", .) %>%        # remove comma
  gsub("\\(D\\)", NA, .) %>%  # replace "(D)" by NA
  as.numeric()                 # change to number

beef0$State <- str_to_title(beef0$State)
beef0$County <- str_to_title(beef0$County)


beef1<-beef0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 5)
    )
  )%>%mutate(County = str_to_title(County))%>%
  dplyr::select(Year, State,County, Value)%>%
    mutate(Value = replace(Value, !is.finite(Value), 0))
  

#reorganize database
beefYear <- beef1 %>%
  pivot_wider(
    names_from = Year,
    values_from = Value)%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


#estimate means
beefYear$Beef_mean<- beefYear$`2022` + beefYear$`2017`/ 2

shp2 <- shp2 %>%#added to the final database 
  left_join(
    beefYear %>%
      dplyr::select(State, County, Beef_mean),
    by = c("NAME_1" = "State",
           "NAME_2" = "County")
  )


##### dairy
#https://quickstats.nass.usda.gov/src/glossary.pdf 

dairy0<- read_csv("folder/CensusMilk_NASSstatequicks_2017_2022.csv")#change by the folder and file of use

#data cleaning
dairy0$Value <- dairy0$Value %>%
  gsub(",", "", .) %>%        
  gsub("\\(D\\)", NA, .) %>%  
  as.numeric()                

dairy0$State <- str_to_title(dairy0$State)
dairy0$County <- str_to_title(dairy0$County)


dairy1<-dairy0%>%
  dplyr::filter(
    sapply(State, function(x)
      any(stringdist(str_to_title(x), str_to_title(states_complete)) < 5)
    )
  )%>%
  dplyr::select(Year, State,County, Value)%>%
    mutate(Value = replace(Value, !is.finite(Value), 0))

#reorganize database
dairyYear <- dairy1 %>%
  pivot_wider(
    names_from = Year,
    values_from = Value)%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


#estimate means
dairyYear$Dairy_mean<- dairyYear$`2022` + dairyYear$`2017`/ 2

shp2 <- shp2 %>%#added to the final database 
  left_join(
    dairyYear %>%
      dplyr::select(State, County, Dairy_mean),
    by = c("NAME_1" = "State",
           "NAME_2" = "County")
  )



##### humans
#https://www.ers.usda.gov/data-products/rural-urban-continuum-codes 

humans<- read_xlsx("folder/Ruralurbancontinuumcodes2023.xlsx")#change by the folder and file of use

#data cleaning
humans1 <- humans %>%
  filter(State %in% states_2code_letter) %>%
  mutate(
    County = str_remove(County_Name, " County"))%>%
  dplyr::select(State, County, Population_2020)
  

humans1 <- humans1 %>%
  mutate(
    COUNTY = ifelse(
      State == "CT",
      sub("^Capitol", "Capital", County),
      County
    )
  )

#Join the planning region for CT
humans1_fixed <- humans1 %>%
  mutate(
    key6 = ifelse(State == "CT", substr(COUNTY, 1, 6), NA)
  ) %>%
  left_join(
    CTregshp %>%
      mutate(key6 = substr(PlanningRe, 1, 6)) %>%
      dplyr::select(key6, NAME_2),
    by = "key6"
  ) %>%
  mutate(
    COUNTY = ifelse(State == "CT" & !is.na(NAME_2), NAME_2, COUNTY)
  ) %>%
  dplyr::select(-key6, -NAME_2)%>%
  dplyr::select(State, COUNTY, Population_2020)


humans1_fixed <- humans1_fixed %>%
  mutate(
    COUNTY = str_to_upper(str_trim(COUNTY)),
    State = str_to_upper(str_trim(State))
  )

shp2 <- shp2 %>%#added to the final database 
  left_join(
    humans1_fixed,
    by = c("states_2code_letter" = "State",
           "NAME_2_upper" = "COUNTY")
  )



##### social index
#https://www.atsdr.cdc.gov/place-health/php/svi/svi-interactive-map.html


social<- read_csv("folder/svi_interactive_map.csv")#change by the folder and file of use

#data cleaning
social1 <- social %>%
  dplyr::filter(ST_ABBR %in% states_2code_letter) %>%
  dplyr::select(ST_ABBR, COUNTY, RPL_THEMES)%>%
  mutate(
    COUNTY = str_remove(COUNTY, " County"))


social1 <- social1 %>%
  mutate(
    COUNTY = ifelse(
      ST_ABBR == "CT",
      sub("^Capitol", "Capital", COUNTY),
      COUNTY
    )
  )

#Join the planning region for CT
social1_fixed <- social1 %>%
  mutate(
    key6 = ifelse(ST_ABBR == "CT", substr(COUNTY, 1, 6), NA)
  ) %>%
  left_join(
    CTregshp %>%
      mutate(key6 = substr(PlanningRe, 1, 6)) %>%
      dplyr::select(key6, NAME_2),
    by = "key6"
  ) %>%
  mutate(
    COUNTY = ifelse(ST_ABBR == "CT" & !is.na(NAME_2), NAME_2, COUNTY)
  ) %>%
  dplyr::select(-key6, -NAME_2)%>%
  dplyr::select(ST_ABBR, COUNTY, RPL_THEMES)


social1_fixed <- social1_fixed %>%
  mutate(
    COUNTY = str_to_upper(str_trim(COUNTY)),
    ST_ABBR = str_to_upper(str_trim(ST_ABBR))
  )


shp2 <- shp2 %>%#added to the final database 
  left_join(
    social1_fixed,
    by = c("states_2code_letter" = "ST_ABBR",
           "NAME_2_upper" = "COUNTY")
  )



```



